Typestate Inference in an Imperative Calculus

### 2012-11-30 formalism/paper.tex
minor clean-ups of proofs of context normalisation, some initial descriptive work of the object constraint solving process.

\newcommand{\Tr}{\mathit{Tr}}
\newcommand{\secref}[1]{Section \ref{#1}}
\newcommand{\figref}[1]{Figure \ref{#1}}
\newcommand{\lstref}[1]{Listing \ref{#1}}
\newcommand{\appref}[1]{Appendix \ref{#1}}
\newcommand{\lemref}[1]{Lemma \ref{#1}}
\newcommand{\corref}[1]{Corollary \ref{#1}}

@@

% unit value
\inferlhs{\unitv} \triangleq \alpha = \unitt \cand \gamma' = \gamma
\\\\

% true value
\inferlhs{\truev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
\\\\

% false value
\inferlhs{\falsev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
\\\\

% object value
\inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@\overline{S}}
\triangleq \\
\exists \alpha_o, \alpha_s, \overline{\alpha_{ij}}, \gamma_{\emptyset} . \\
\gamma_{\emptyset} = \emptyset
\cand
\alpha = \alpha_o@\alpha_s
\cand
\overline{\inferrule{\gamma_{\emptyset}}{v_{ij}}{\alpha_{ij}}{\gamma_{\emptyset}}}
\cand
\\
\gamma = \gamma'
\cand
\alpha_o @ \alpha_s = \{ \overline{ S_i \{ \overline{ m_{ij} : \alpha_{ij} >> S_{ij} } \} }\} @ \overline{S}
\\\\

% function value
\inferlhs{\lambda ( \overline{x_i : T_i \gg T_i'} ) . t} \triangleq \\
\exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_i, \alpha_i'}, \alpha_r . \\
\alpha = (\overline{\alpha_i \gg \alpha_i'}) \rightarrow \alpha_r
\cand
\gamma = \gamma'
\cand
\gamma_r = \emptyset
\cand \\
\gamma_{in} = \gamma_r, \overline{x_i : \alpha_i}
\cand
\overline{x_i : \alpha_i' \in \gamma_{out}}
\cand
\\
\inferrule{\gamma_{in}}{t}{\alpha_r}{\gamma_{out}}
\\
\cand
\overline{T_i <: \alpha_i}
\cand
\overline{\alpha_i' <: T_i'}
\\\\

% let binding
\inferlhs{\lett{x}{t'}{t}} \triangleq \\
\exists \gamma_v, \gamma_{in}, \gamma_{out}, \alpha' . \\
\gamma_{in} = \gamma_v, x : \alpha'
\cand
\gamma' = \gamma_{out} - \{ x \}
\cand \\
\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}
\cand
\inferrule{\gamma_{in}}{t}{\alpha}{\gamma_{out}}
\\\\

% sequence
\inferlhs{t ; t'} \triangleq \\
\exists \gamma'', \alpha' . \\
\inferrule{\gamma}{t}{\alpha'}{\gamma''}
\cand
\inferrule{\gamma''}{t'}{\alpha}{\gamma'}
\\\\

% function call
\inferlhs{x ( \overline{x_i} )} \triangleq \\
\exists \overline{\alpha_i}, \overline{\alpha_i'}, \overline{\alpha_i''} . \\
\overline{x_i : \alpha_i \in \gamma}
\cand 
x : ( \overline{\alpha_i' \gg \alpha_i''} ) \rightarrow \alpha \in \gamma
\cand
\\
\gamma' = \gamma [ \overline{x_i \mapsto \remapapp{\alpha_i}{\alpha_i'}{\alpha_i''} } ]
\\\\

% method call
\inferlhs{x.m} \triangleq \\
\exists \alpha_o, \alpha_s . \\
x : \alpha_o@\alpha_s \in \gamma
\cand \\
\gamma' = \gamma[ x \mapsto \remapapp{\alpha_o@\alpha_s}{O@S}{O@S'} ]
\\
\mathbf{where}\; O = \{ S \{ m : \alpha \Rightarrow S' \} S' \{ \} \}
\\\\

% if-then-else
\inferlhs{\ift{t_c}{t_t}{t_f}} \triangleq \\
\exists \alpha_c, \alpha_t, \alpha_f, \gamma_c, \gamma_t, \gamma_f . \\
\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c}
\cand
\inferrule{\gamma_c}{t_t}{\alpha_t}{\gamma_t}
\cand 
\\
\inferrule{\gamma_c}{t_f}{\alpha_f}{\gamma_f}
\cand
\alpha_c = \boolt
\cand
\\
\alpha = \alpha_t \sqcupplus \alpha_f
\cand
\gamma' = \gamma_t \sqcup \gamma_f
\\\\

@@

Additionally, we have {\bf variable usage constraints} of form 
$x : \tinf \in \gamma$, which are generated for each location in the
program where a variable is used and what type that usage implies. These
are used to generate additional type variable
constraints by matching the asserted type $\tinf$ with what is known
about the type of $x$ in $\gamma$.

\item {\bf Type expressions constraints} --- these express 
constraints between two or three types: either two types are equal,
or one type is the join of two others. The latter form of type constraint
is generated when dealing with if-then-else terms.
\end{itemize}

@@

A constraint generation expression $\inferlhs{t}$, which relates the term
to an input context variable, a type variable and an output context variable, 
is expanded using the rules shown in \figref{fig:constraintgen}.
The constraint generation process for a root term $t$ begins with
$\exists \gamma, \gamma', \alpha. \free{\gamma} \cand \inferlhs{t}$.

@@

\begin{itemize}
\item All value terms produce a $\gamma = \gamma'$ constraint as values do
not change the context. Additionally, a type variable constraint of form
$\alpha = \tinf$ where $\tinf$ is of the appropriate type to match
the value, with type variables in all positions relating to parameters and return
values.
\item The context variable $\gamma_a$ used as the input context for
function bodies has an empty root $\gamma_r$, to allow for detection of 
used unbound variables within
the body. This matches the use of a context in type inference
composed only of the parameters, and prevents implicit binding.
\item Let-bind terms generates a new context variable $\gamma_v$ for the
output context of the value term. This is extended by $\gamma_{b1}$, the
input context for the body term, with variable $x$ of the type inferred
for the value. The output context of the whole let-bind term removes
$x$ from the output context $\gamma_{b2}$ of the body term.
\item Update terms generate replacement constraints based on the inferred
type $\alpha'$ of the right-hand side of the expression.
\item Sequencing terms relate the output of the left term to the input of the
right term through a new context variable $\gamma''$. The type variable
$\alpha'$ must be generated to capture the type of the left term for
consistency, but it is not needed.
\item Function call terms generate context variable constraints to relate the
usage of the variables to whatever information has been inferred about the
parameter types of the function. All variables $\overline{x_i}$ used as 
parameters in the function call are replaced in the output context with
new type bindings to $\overline{\alpha_i'}$, which represents the effect
of the function.
\item Method call terms generate type variables to capture that $x$ must
be an object, that the method $m$ must exist in the current state $\alpha_s$
and that after the method call $x$ must be in the new state $\alpha_s'$.
\end{itemize}

@@

Constraint solving is performed in phases. First, generated context constraints
for a term $t$ aresolved, which determines the contents of each context associated to
each context variable in the typing derivation of $t$, and produces additional
type equality constraints (of form $\mathbb{T} = \mathbb{T}'$) based on
observed usages of variables. If a variable either used in a context where
it is unbound (the variable is {\it free} in $t$) or it is rebound
(a subterm within $t$ is of form $let x = ... in t'$ where a subterm of $t'$
is also of form $let x = ... in t''$), then no principal typing exists and
the phase reports the offending variable.

@@

Solving context constraints involves {\it normalising} the constraints. 
Normalisation means that every context variable will have a single 
defining constraint of form $\gamma = \emptyset$, $\free{\gamma}$ or
$\gamma = \gamma', \overline{x_i : \tinf_i}$, where $\gamma'$ is
either empty or free. A context $\Gamma$ for type checking can be easily
constructed from constraints in this form, as once the type variable
constraints are solved we can say $\Gamma = \overline{x_i : T_i}$ for
every $\gamma = \gamma', \overline{x_i : \tinf_i}$ where $T_i$ is the
solved form of $\tinf_i$.

We say that a constraint $C$ is {\it normalisable} if, by
applying the rules in \figref{fig:contextsolve}, it can be transformed to
a constraint $C'$ which is normalised or the rules produce $\mathbf{bound}\:x$
or $\mathbf{unbound}\:x$ for some variable $x$.

@@

\begin{itemize}
\item \textsc{ceq} --- Where $\gamma$ is directly equivalent
to $\gamma'$, which is defined in terms of extension of
$\gamma''$, rewrite the constraints such that
$\gamma$ directly copies the extension of $\gamma''$.
\item \textsc{crem} --- Where $\gamma$ removes a variable $x$ from $\gamma'$,
which is defined in terms of extension of $\gamma''$, rewrite the
constraints such that $\gamma$ copies the extension of $\gamma''$ with
the variable $x$ removed from the extension. There is no need for an
error-producing variant of \textsc{crem}, where the variable $x$ is missing, 
as context removal constraints are always generated in a pair with a context 
extension constraint for let binding.
\item \textsc{cext} --- Where $\gamma$ extends $\gamma'$ with
a variable set $\overline{y_j}$, and $\gamma'$ is defined in terms of extension of
$\gamma''$ with a disjoint set of variables in its extension set, rewrite the
constraints such that $\gamma$'s extension set includes both the $y_j$
definitions and $x_j$ definitions.
\item $\textsc{cext}_2$ --- Where $\gamma$ extends $\gamma'$ with
a new variable $x$, and $\gamma'$ is defined in terms of extension of
some $\gamma''$ and the extension set already contains $x$, this
represents an attempt to redeclare $x$. As such, generate an error
$\mathbf{bound}\:x$, terminating the constraint solving process.
\item \textsc{crep} --- where $\gamma$ replaces the type of some
variable $x$ in $\gamma'$, where $\gamma$ is defined in terms of extension
of $\gamma''$ and $x$ is in the extension set, rewrite the constraints
such that $\gamma$ copies the extension set of $\gamma'$ with $x$
replaced as specified.
\item $\textsc{crep}_2$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of extension of $\gamma''$
which is free and $x$ is not in the extension set, $x$ is a
free variable in the term being processed. To introduce $x$ in such a way that
it will be included in all contexts, $\gamma''$ is redefined in
terms of extension of a new context $\gamma_r$ with variable $x$
in the extension set mapped to a new type variable $\alpha$. Following
this, \textsc{cext} can be applied to $\gamma'$, then \textsc{crep} can
be applied to $\gamma$.
\item $\textsc{crep}_3$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of an empty root $\gamma''$,
and $x$ is not in $\gamma$'s extension set, this indicates the usage
of an unbound variable within a function body. As such, generate an
error $\mathbf{unbound}\:x$ indicating that x is unbound within this
context, terminating the contraint solving process.
\item \textsc{cvar} --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of
extension of $\gamma_r$ which is free, and $x$ is not in the extension
set of $\gamma$, $x$ is a free variable in the term being processed. 
Similar to $\textsc{crep}_2$, redefine $\gamma_r$
in terms of extension of a new root $\gamma_r'$, with $x$ in
the extension set with the observed type $\tinf$.
\item $\textsc{cvar}_2$ --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of $\gamma'$
which is empty, and $x$ is not in the extension set of $\gamma$, 
$x$ is an unbound variable within a function body.
\end{itemize}

@@

The way in which the context constraints are generated guarantees
that the dependencies between contexts form a directed acyclic graph. For
this reason, we believe that that the first phase will always terminate,
but this has not yet been proven.

\begin{center}
\begin{tabular}{c}
	\infer[\textsc{c\_eq}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
	{\gamma = \gamma' \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
\\\\
	\infer[\textsc{c\_eq\_2}]
	{\gamma = \gamma', \emptyset}
	{C \cand \gamma = \gamma' & \free{\gamma'} \in C \vee \gamma' = \emptyset \in C}
\\\\
	\infer[\textsc{c\_rem}]
	{
	\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
	}
	{
	\gamma = \gamma' - \{x\} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
	}
\\\\
	\infer[\textsc{c\_ext}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i}, \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
	{
		\gamma = \gamma', \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
		& \overline{y_j} \cap \overline{x_i} = \emptyset
	}
\\\\
	\infer[\textsc{c\_ext\_2}]
	{\mathbf{bound}\:x}
	{
		\gamma = \gamma', \overline{y_j : \tinf_j}, x : \tinf \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
		& x \in \overline{x_i}
	}
\\\\
	\infer[\textsc{c\_rep}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i}, x : \tinf \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
	{\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
\\\\
	\infer[\textsc{c\_rep\_2}]
	{
	\begin{array}{c}
	\exists \alpha . 
	\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i} \cand
	\\
	x : \alpha \in \gamma''
	\end{array}
	}
	{
	\begin{array}{c}
	\begin{array}{cc}
	x \notin \overline{x_i}
	&
	\fresh{\alpha}
	\end{array}
	\\
	\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}
	\end{array}
	}
\\\\
	\infer[\textsc{crep\_3}]
	{C \cand \gamma = \gamma'[x \mapsto \tinf] \cand 
	 x : \alpha \in \gamma'}
	{
	C \cand \gamma = \gamma'[x \mapsto \tinf]
	&
	\free{\gamma'} \in C \vee \gamma' = \emptyset \in C
	}
\\\\
	\infer[\textsc{cvar\_1}]
	{\exists \gamma' . \free{\gamma'} \cand \gamma = \gamma', x : \tinf}
	{\free{\gamma} \cand x : \tinf \in \gamma}
\\\\
	\infer[\textsc{cvar\_2}]
	{\mathbf{unbound}\:x}
	{\gamma = \emptyset \cand x : \tinf \in \gamma}
\\\\
	\infer[\textsc{cvar\_3}]
	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma'}
	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma
	&
	x \notin \overline{x_i}
	}
\\\\
	\infer[\textsc{c\_join}]
	{
	\begin{array}{c}
	\gamma = \overline{x_i : \tinf_i''} \cand 
	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
	\\
	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'} \cand
	\overline{\tinf_i'' = \tinf_i \sqcup \tinf_i'}
	\end{array}
	}
	{
	\begin{array}{c}
	\gamma = \gamma' \sqcup \gamma'' \cand 
	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
	\\
	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'}
	\end{array}
	}
\\\\
	\infer[\textsc{cerr\_1}]
	{\mathbf{unbound}\:x}
	{C \cand \mathbf{unbound}\:x}
\\\\
	\infer[\textsc{cerr\_2}]
	{\mathbf{bound}\:x}
	{C \cand \mathbf{bound}\:x}
\end{tabular}
\end{center}
\caption{\label{fig:contextsolve} Context normalisation}
\end{figure}

@@

\subsection{Object variable constraint solving}

>>>>

\newcommand{\Tr}{\mathit{Tr}}
\newcommand{\secref}[1]{Section \ref{#1}}
\newcommand{\figref}[1]{Figure \ref{#1}}
\newcommand{\lstref}[1]{Listing \ref{#1}}
\newcommand{\appref}[1]{Appendix \ref{#1}}
\newcommand{\thmref}[1]{Theorem \ref{#1}}
\newcommand{\lemref}[1]{Lemma \ref{#1}}
\newcommand{\corref}[1]{Corollary \ref{#1}}

@@

% unit value
\inferlhs{\unitv} \coloneqq \alpha = \unitt \cand \gamma' = \gamma
\\\\

% true value
\inferlhs{\truev} \coloneqq \alpha = \boolt \cand \gamma' = \gamma 
\\\\

% false value
\inferlhs{\falsev} \coloneqq \alpha = \boolt \cand \gamma' = \gamma 
\\\\

% object value
\inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@\overline{S}}
\coloneqq \\
\exists \alpha_o, \alpha_s, \overline{\alpha_{ij}}, \gamma_{\emptyset} . \\
\gamma_{\emptyset} = \emptyset
\cand
\alpha = \alpha_o@\alpha_s
\cand
\overline{\inferrule{\gamma_{\emptyset}}{v_{ij}}{\alpha_{ij}}{\gamma_{\emptyset}}}
\cand
\\
\gamma = \gamma'
\cand
\alpha_o @ \alpha_s = \{ \overline{ S_i \{ \overline{ m_{ij} : \alpha_{ij} >> S_{ij} } \} }\} @ \overline{S}
\\\\

% function value
\inferlhs{\lambda ( \overline{x_i : T_i \gg T_i'} ) . t} \coloneqq \\
\exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_i, \alpha_i'}, \alpha_r . \\
\alpha = (\overline{\alpha_i \gg \alpha_i'}) \rightarrow \alpha_r
\cand
\gamma = \gamma'
\cand
\gamma_r = \emptyset
\cand \\
\gamma_{in} = \gamma_r, \overline{x_i : \alpha_i}
\cand
\overline{x_i : \alpha_i' \in \gamma_{out}}
\cand
\\
\inferrule{\gamma_{in}}{t}{\alpha_r}{\gamma_{out}}
\\
\cand
\overline{T_i <: \alpha_i}
\cand
\overline{\alpha_i' <: T_i'}
\\\\

% let binding
\inferlhs{\lett{x}{t'}{t}} \coloneqq \\
\exists \gamma_v, \gamma_{in}, \gamma_{out}, \alpha' . \\
\gamma_{in} = \gamma_v, x : \alpha'
\cand
\gamma' = \gamma_{out} - \{ x \}
\cand \\
\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}
\cand
\inferrule{\gamma_{in}}{t}{\alpha}{\gamma_{out}}
\\\\

% sequence
\inferlhs{t ; t'} \coloneqq \\
\exists \gamma'', \alpha' . \\
\inferrule{\gamma}{t}{\alpha'}{\gamma''}
\cand
\inferrule{\gamma''}{t'}{\alpha}{\gamma'}
\\\\

% function call
\inferlhs{x ( \overline{x_i} )} \coloneqq \\
\exists \overline{\alpha_i}, \overline{\alpha_i'}, \overline{\alpha_i''} . \\
\overline{x_i : \alpha_i \in \gamma}
\cand 
x : ( \overline{\alpha_i' \gg \alpha_i''} ) \rightarrow \alpha \in \gamma
\cand
\\
\gamma' = \gamma [ \overline{x_i \mapsto \remapapp{\alpha_i}{\alpha_i'}{\alpha_i''} } ]
\\\\

% method call
\inferlhs{x.m} \coloneqq \\
\exists \alpha_o, \alpha_s . \\
x : \alpha_o@\alpha_s \in \gamma
\cand \\
\gamma' = \gamma[ x \mapsto \remapapp{\alpha_o@\alpha_s}{O@S}{O@S'} ]
\\
\mathbf{where}\; O = \{ S \{ m : \alpha \Rightarrow S' \} S' \{ \} \}
\\\\

% if-then-else
\inferlhs{\ift{t_c}{t_t}{t_f}} \coloneqq \\
\exists \alpha_c, \alpha_t, \alpha_f, \gamma_c, \gamma_t, \gamma_f . \\
\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c}
\cand
\inferrule{\gamma_c}{t_t}{\alpha_t}{\gamma_t}
\cand 
\\
\inferrule{\gamma_c}{t_f}{\alpha_f}{\gamma_f}
\cand
\alpha_c = \boolt
\cand
\\
\alpha = \alpha_t \sqcupplus \alpha_f
\cand
\gamma' = \gamma_t \sqcup \gamma_f
\\\\

@@

Additionally, we have {\it variable usage constraints} of form 
$x : \tinf \in \gamma$, which are generated for each location in the
program where a variable is used and what type that usage implies. These
are used to generate additional type variable
constraints by matching the asserted type $\tinf$ with what is known
about the type of $x$ in $\gamma$.

\item {\bf Type expressions constraints} --- these express 
relationships between types. Either two types are equal ($\tinf = \tinf'$),
one type is a subtype of another ($\tinf <: \tinf'$) or one type
is the join of two others ($\tinf \sqcup \tinf'$). As a special case of
the latter, if it is known that the two types are derived from
interactions with the same value, this is written as $\tinf \veebar \tinf'$.
This additional information is relevant when solving object type constraints
in particular, and is discussed later.
\end{itemize}

@@

A constraint generation expression $\inferlhs{t}$, which relates the term
to an input context variable, a type variable and an output context variable, 
is expanded using the rules shown in \figref{fig:constraintgen}.
The constraint generation process for a root term is defined
by the function $gen(t) = \exists \gamma, \gamma', \alpha. \free{\gamma} \cand \inferlhs{t}$.

@@

\begin{itemize}
\item All value terms produce a $\gamma = \gamma'$ constraint as values do
not change the context. Additionally, a type variable constraint of form
$\alpha = \tinf$ is generated where $\tinf$ is of the appropriate type to 
match the value. In the case of functions or object values, this type will
contain additional type variables for parameters and return values, with
the intention that these be solved to validate user type annotations.
\item The context variable $\gamma_{in}$ used as the input context for
function bodies has an empty root $\gamma_r$. This allows for the detection of 
used unbound variables within
the body term. The domain of the input context is strictly composed of the 
parameters, preventing implicit binding, in order to match the properties
of the typing judgement for functions.
\item Let-bind terms generates a new context variable $\gamma_v$ for the
output context of the value term. This is extended by $\gamma_{in}$, the
input context for the body term, with variable $x$ of the type inferred
for the value. The output context $\gamma'$ of the whole let-bind term removes
$x$ from the output context $\gamma_{out}$ of the body term.
\item Sequencing terms relate the output of the left term to the input of the
right term through a new context variable $\gamma''$. The type variable
$\alpha'$ is generated to infer the type of the left term, though this is not needed beyond confirming that it does in fact have a type.
\item Function call terms generate context variable constraints to relate the
usage of the variables to whatever information has been inferred about the
parameter types of the function. All variables $\overline{x_i}$ used as 
parameters in the function call are replaced in the output context with
new type bindings to $\overline{\alpha_i'}$, which represents the effect
of the function.
\item Method call terms generate type variables to capture that $x$ must
be an object, that the method $m$ must exist in the current state $\alpha_s$
and that after the method call $x$ must be in the new state $\alpha_s'$.
\item If-then-else terms relate the context variables used for the
condition, true branch and false branch as expected. It is asserted that the
type of the condition must be a boolean. The output context
variable $\gamma'$ is expressed as the join of the outputs for the true
and false branches.
\end{itemize}

@@

Constraint solving is performed in phases. First, generated context constraints
for a term $t$ are solved, which determines the contents of each context associated to
each context variable in the typing derivation of $t$, and produces additional
type equality constraints (of form $\mathbb{T} = \mathbb{T}'$) based on
observed usages of variables. If a variable either used in a context where
it is unbound (the variable is {\it free} in $t$) or it is rebound
(a subterm within $t$ is of form $let x = ... in t'$ where a subterm of $t'$
is also of form $let x = ... in t''$), then no principal typing exists and
the phase reports the offending variable.

@@

Solving context constraints involves {\it normalising} the constraints.
A {\it normalised} constraint is one in which
context variable has a single 
defining constraint of form $\gamma = \emptyset$, $\free{\gamma}$ or
$\gamma = \gamma', \overline{x_i : \tinf_i}$, where $\gamma'$ is
either empty or free. A context $\Gamma$ for type checking can be easily
constructed from constraints in this form, as once the type variable
constraints are solved we can say $\Gamma = \overline{x_i : T_i}$ for
every $\gamma = \gamma', \overline{x_i : \tinf_i}$ where $T_i$ is the
solved form of $\tinf_i$, or $\Gamma = \emptyset$ for every $\free{\gamma}$
or $\gamma = \emptyset$.

We say that a constraint $C$ is {\it normalisable} if, by
applying the rules in \figref{fig:contextsolve}, it can be transformed to
a constraint $C'$ which is normalised, or the rules produce $\mathbf{bound}\:x$
or $\mathbf{unbound}\:x$ for some variable $x$.

\begin{center}
\begin{tabular}{c}
	\infer[\textsc{c\_eq}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
	{\gamma = \gamma' \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
\\\\
	\infer[\textsc{c\_eq\_2}]
	{\gamma = \gamma', \emptyset}
	{C \cand \gamma = \gamma' & \free{\gamma'} \in C \vee \gamma' = \emptyset \in C}
\\\\
	\infer[\textsc{c\_rem}]
	{
	\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
	}
	{
	\gamma = \gamma' - \{x\} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
	}
\\\\
	\infer[\textsc{c\_ext}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i}, \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
	{
		\gamma = \gamma', \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
		& \overline{y_j} \cap \overline{x_i} = \emptyset
	}
\\\\
	\infer[\textsc{c\_ext\_2}]
	{\mathbf{bound}\:x}
	{
		\gamma = \gamma', \overline{y_j : \tinf_j}, x : \tinf \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
		& x \in \overline{x_i}
	}
\\\\
	\infer[\textsc{c\_rep}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i}, x : \tinf \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
	{\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
\\\\
	\infer[\textsc{c\_rep\_2}]
	{
	\begin{array}{c}
	\exists \alpha . 
	\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i} \cand
	\\
	x : \alpha \in \gamma''
	\end{array}
	}
	{
	\begin{array}{c}
	\begin{array}{cc}
	x \notin \overline{x_i}
	&
	\fresh{\alpha}
	\end{array}
	\\
	\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}
	\end{array}
	}
\\\\
	\infer[\textsc{crep\_3}]
	{C \cand \gamma = \gamma'[x \mapsto \tinf] \cand 
	 x : \alpha \in \gamma'}
	{
	C \cand \gamma = \gamma'[x \mapsto \tinf]
	&
	\free{\gamma'} \in C \vee \gamma' = \emptyset \in C
	}
\\\\
	\infer[\textsc{cvar\_1}]
	{\exists \gamma' . \free{\gamma'} \cand \gamma = \gamma', x : \tinf}
	{\free{\gamma} \cand x : \tinf \in \gamma}
\\\\
	\infer[\textsc{cvar\_2}]
	{\mathbf{unbound}\:x}
	{\gamma = \emptyset \cand x : \tinf \in \gamma}
\\\\
	\infer[\textsc{cvar\_3}]
	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma'}
	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma
	&
	x \notin \overline{x_i}
	}
\\\\
\infer[\textsc{t\_var}]
	{
	\tinf' <: \tinf \cand 
	\gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'
	}
	{
	x : \tinf \in \gamma \cand 
	\gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
\\\\
	\infer[\textsc{c\_join}]
	{
	\begin{array}{c}
	\gamma = \overline{x_i : \tinf_i''} \cand 
	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
	\\
	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'} \cand
	\overline{\tinf_i'' = \tinf_i \sqcup \tinf_i'}
	\end{array}
	}
	{
	\begin{array}{c}
	\gamma = \gamma' \sqcup \gamma'' \cand 
	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
	\\
	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'}
	\end{array}
	}
\\\\
	\infer[\textsc{cerr\_1}]
	{\mathbf{unbound}\:x}
	{C \cand \mathbf{unbound}\:x}
\\\\
	\infer[\textsc{cerr\_2}]
	{\mathbf{bound}\:x}
	{C \cand \mathbf{bound}\:x}
\end{tabular}
\end{center}
\caption{\label{fig:contextsolve} Context normalisation}
\end{figure}

@@

\begin{itemize}
\item $\textsc{c\_eq}$ --- Where $\gamma$ is directly equivalent
to $\gamma'$, which is defined in terms of extension of
$\gamma''$, rewrite the constraints such that
$\gamma$ directly copies the definition of $\gamma'$.
\item $\textsc{c\_eq\_2}$ --- Where $\gamma$ is directly equivalent
to $\gamma'$, which is defined as either $\free{\gamma}$ or 
$\gamma = \emptyset$, $\gamma'$ is defined to be an empty extension of 
$\gamma'$.

\item $\textsc{c\_rem}$ --- Where $\gamma$ removes a variable $x$ from $\gamma'$,
which is defined in terms of extension of $\gamma''$ that is known to have
$x$ in its domain, rewrite the
constraints such that $\gamma$ copies the definition $\gamma''$ with
the variable $x$ removed from the extension. There is no need for an
error-producing variant of \textsc{c\_rem}, where the variable $x$ is missing, 
as context removal constraints are always generated in a pair with a context 
extension constraint for let binding.

\item $\textsc{c\_ext}$ --- Where $\gamma$ extends $\gamma'$ with
a variable set $\overline{y_j}$, and $\gamma'$ is defined in terms of extension of
$\gamma''$ with a disjoint set of variables in its extension set, rewrite the
constraint such that $\gamma$'s extension set includes both the $y_j$
definitions and $x_j$ definitions.

\item $\textsc{c\_ext}_2$ --- Where $\gamma$ extends $\gamma'$ with
a new variable $x$, and $\gamma'$ is defined in terms of extension of
some $\gamma''$ and the extension set already contains $x$, this
represents an attempt to redeclare $x$. As such, generate an error
$\mathbf{bound}\:x$, terminating the constraint solving process.

\item $\textsc{c\_rep}$ --- where $\gamma$ replaces the type of some
variable $x$ in $\gamma'$, where $\gamma$ is defined in terms of extension
of $\gamma''$ and $x$ is in the extension set, rewrite the constraints
such that $\gamma$ copies the extension set of $\gamma'$ with $x$
replaced as specified.

\item $\textsc{c\_rep}_2$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of extension of $\gamma''$
which is free and $x$ is not in the extension set, $x$ is a
free variable in the term being processed. To introduce $x$ in such a way that
it will be included in all contexts, a variable usage constraint
$x : \alpha \in \gamma''$ is introduced, with the intention that this be
solved by $c\_var\_1$ or $c\_var\_2$ first before the replacement can be
solved using $c\_rep\_1$.

\item $\textsc{crep}_3$ --- this performs the same function as
$c\_rep\_2$, but for the case where $\gamma$ is directly defined as
a replacement on a free or empty context variable $\gamma'$.

\item $\textsc{c\_var\_1}$ --- where $x$ is asserted to exist in a context
variable $\gamma$, which is defined as $\free{\gamma}$, the definition
of $\gamma$ is replaced with one which extends a new free context variable
$\gamma'$ with the variable $x$. This is intended to also have the effect
that all other context variables which are defined as extensions of 
$\gamma$ will have to be updated using \textsc{c\_ext}, or an error
generated by \textsc{c\_ext} if they already contain $x$.

\item $\textsc{c\_var\_2}$ --- where $x$ is asserted to exist in a
context variable $\gamma$, which is defined as $\gamma = \emptyset$, this
represents the attempt to use an unbound variable in a context which does
not permit free variables (such as in the body of a function). As such,
an $\mathbf{unbound}\:x$ error is generated.

\item $\textsc{c\_var\_3}$ --- where $x$ is asserted to exist
in a context variable $\gamma$, which is defined in terms of
extension of $\gamma'$ which does not include $x$, the assertion
can be rewritten as an assertion that $x$ must exist within $\gamma'$.
This is intended to propagate to a root context variable and solved
by \textsc{c\_var\_1} or \textsc{c\_var\_2}.

\item $\textsc{c\_join}$ --- where a context variable is defined as the
join of two other context variables $\gamma'$ and $\gamma''$, which are in turn defined as extensions
of the same context variable $\gamma_r$, and which contain the same variables,
the join can be solved by generating a new set of type variables, one for
each $x_i$, defined as the join of the types known for $x_i$ in $\gamma'$
and $\gamma''$.
\end{itemize}

@@

Given an arbitrary term $t$, the context constraints generated by $t$
are normalisable (see \appref{app:ctx_norm}).
This is less general than stating that the context constraint solving rules
can solve any conjunction of context constraints, but is sufficient for
the purposes of this language.

@@

Solving a set of type variable constraints consists of demonstrating that
there exists at least one substitution $\sigma$ for all type variables that 
turns all constraints into tautologies.
